---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: TRUE
---
```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
```

## Loading and preprocessing the data

```{r}
data <- read.csv("activity.csv")

data <- data %>%
  mutate(steps = as.numeric(steps),
         interval = as.numeric(interval),
         date = as.Date(date))
```

## What is mean total number of steps taken per day?

##### 1. Calculate the total number of steps taken per day
```{r}
steps_day <- data %>%
  filter(!is.na(data)) %>%
  group_by(date) %>%
  summarise(across(steps, sum, na.rm = TRUE))
steps_day
```

##### 2. Make a histogram of the total number of steps taken each day
```{r}
hist(steps_day$steps,
     main = "Total number of steps taken per day",
     xlab = "Number of steps",
     col = "lightblue1",
     labels = T,
     breaks = seq(0,22000, by = 2000),
     xaxp = c(0, 22000, 11),
     ylim = c(0, 18),
     yaxp = c(0, 16, 8))

abline(h = seq(0, 16, 2), col = "gray", lty = "dotted")
```

##### 3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
steps_mean <- mean(steps_day$steps)
steps_median <- median(steps_day$steps)
steps_mean
steps_median
```

## What is the average daily activity pattern?
##### 1. Make a time series plot
```{r}
steps_intv <- data %>% 
  group_by(interval) %>%
  summarise(steps = mean(steps, na.rm =TRUE))

ggplot(steps_intv,aes(interval, steps)) +
  geom_line(size = 0.5, col = "blue") +
  xlab("5-minute interval") +
  ylab("average number of steps")
```

##### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
steps_intv %>%
  arrange(desc(steps)) %>%
  slice(1)
```

## Imputing missing values

##### 1. Total number of missing values

```{r}
na <- !complete.cases(data)
sum(na == TRUE)

```

##### 2.-3. Imputing all of the missing values and create new dataset

```{r}
# Impute NA's with mean by interval
data.impute <- data %>%
  group_by(interval) %>%
  mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = T), steps))

```

##### 4. Make a histogram of the total number of steps taken each day 

```{r}
steps_day_imp <- data.impute %>%
  group_by(date) %>%
  summarise(steps = sum(steps))
head(steps_day_imp)

hist(steps_day_imp$steps,
     main = "Total number of steps taken per day (imputed)",
     xlab = "Number of steps",
     col = "lightblue1",
     labels = T,
     breaks = seq(0,22000, by = 2000),
     xaxp = c(0, 22000, 11),
     ylim = c(0, 25),
     yaxp = c(0, 24, 8))

abline(h = seq(0, 24, 2), col = "gray", lty = "dotted")

# Calculate mean and median of imputed data
steps_imp_mean <- mean(steps_day_imp$steps)
steps_imp_median <- median(steps_day_imp$steps)

# Compare mean and median of imputed data with data excluding NA's
name <- c("mean","median")
steps_excl_na <- c(steps_mean, steps_median)
steps_imputed <- c(steps_imp_mean, steps_imp_median)

data.frame(name, steps_excl_na, steps_imputed)

```
###### Mean and median of imputed data are equal. Compared to data with with excluding NA's, just the medians slightly deviate, the means are equal.

## Are there differences in activity patterns between weekdays and weekends?


##### 1. Create new factor variable with "weekend" and "weekday"

```{r}
steps_intv_weektype <- data.impute %>%
  mutate(type_of_day = ifelse(weekdays(date) == "Samstag" | weekdays(date) == "Sonntag", "weekend", "weekday")) %>%
  group_by(interval, type_of_day) %>%
  summarise(steps = mean(steps, na.rm =TRUE))
steps_intv_weektype
```

##### 2. Make a panel plot
```{r}
ggplot(steps_intv_weektype, aes(interval, steps, color = type_of_day)) +
  geom_line(size = 0.5) +
  scale_colour_discrete(name = "Type of Day") +
  xlab("5-minute interval") +
  ylab("average number of steps") +
  facet_wrap(~ type_of_day, ncol = 1, nrow = 2)
```

